\pkg{evtree}: Evolutionary Learning of Globally Optimal Classification and Regression Trees in \proglang{R}

### 2011-09-18 pkg/evtree/inst/doc/evtree.Rnw
started new 'implementation and application in practice section'
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@241 255f4249-2263-4417-b1ea-8498779e27a7

Based on the input data $X$ and the paramters $\theta$, $f(X,\theta)$ gives predictions for the output $Y$. $loss$ is a suitable loss function for the domain of $Y$. Commonly, the misclassification error for classification problems, or the squared error for regression problems, is used. $comp$ quantifies the complexity of trees and penalizes for more complex models in the tree selection process. In CART, the complexity is a non-negative function of the terminal-nodes. Note that finding $\hat{\theta}$ requires a search over all $\Theta_m$.

@@

\section{Implementation}
\label{impl}

\code{evtree} provides an efficient implementation of an evolutionary algorithm that builds classification trees in \proglang{R}. CPU- and memory- intensive tasks are fully computed in \proglang{C++}, while the user interfaces and plot functions are written in \proglang{R}. The \code{.C()} interface \citep{chambers2008software} was used to pass arguments between the two languages. \code{evtree} depends on the \pkg{partykit} package  \citep{hothorn2009partykit}, which provides an infrastructure for representing, summarizing, and visualizing tree-structured models. The supplied functionality includes standard \code{print()}, \code{summary()}, and \code{plot()} functions to display trees and a \code{predict()} function to predict the class labels of new data. The arguments of  \code{evtree()} are
%
\begin{Code}
  evtree(formula, data = list(), weights = NULL, subset = NULL,
    control = evtree.control(...), ...)
\end{Code}
%
where \code{formula} and \code{data} are a standard \proglang{R} formula of type \code{y ~ x1 + x2} with the respective data. \code{weights} is an optional vector specifying weights to be used in the fitting process. The option vector subset specifies a subset of observations to be used in the fitting process. \code{control} comprises a list of control parameters
%
\begin{Code}
  evtree.control(minbucket = 7L, minsplit = 20L, maxdepth = 9L, 
    niterations = 10000L, ntrees = 100L, alpha = 1,
    operatorprob = list(pmutatemajor = 0.2, pmutateminor = 0.2,
      pcrossover = 0.2, psplit = 0.2, pprune = 0.2),
    seed = NULL, ...)
\end{Code}
%
where the parameters \code{minbucket}, \code{minsplit}, and \code{maxdepth} constrain the solution to a minimum number of observations in each terminal node, a minimum number of observation in each internal node and a maximum tree depth. Note that the memory requirements increase by the square of the maximum tree depth. Parameter \code{alpha} regulates the complexity part of the evaluation function and refers to $\alpha$ in Equation~\ref{eq:classification} for classification trees and Equation~\ref{eq:regression} for regression trees. \code{niterations} and \code{ntrees} specify the maximum number of iterations and the number of trees in the population. With the argument \code{operatorprob}, user-specified probabilities for the variation operators can be defined. Argument \code{seed} is an optional seed for the random number generator, which allows models to be reproduced by re-using the same seed. If not specified, the random number generator is initialized by the system time. In this case, iterative calls of \code{evtree} will possibly result in different models.
%
<<iris-tree>>=
library("evtree")
data("iris", package = "datasets")
ev <- evtree(Species ~ ., iris, control = evtree.control(seed = 1000))
@
%
The fitted model can be examined using the \code{print()} and \code{plot()} functions. 
%
<<iris-print>>=
ev
plot(ev)
@

\setkeys{Gin}{width=1\textwidth}
\begin{figure}[t!]
\centering
<<iris-plot, fig=TRUE, echo=FALSE, width=13.2, height=8>>=
plot(ev)
@
\caption{\label{fig:example} Example of a tree plotted by the \code{plot()} function.}
\end{figure}

The resulting plot is displayed in Figure~\ref{fig:example}. \code{predict()} can be used for the prediction of the response.
%
<<iris-predict>>=
predict(ev, newdata = iris[c(10, 35, 60, 85, 110, 135), -5])
@

>>>>

Based on the input data $X$ and the parameters $\theta$, $f(X,\theta)$ gives predictions for the output $Y$. $loss$ is a suitable loss function for the domain of $Y$. Commonly, the misclassification error for classification problems, or the squared error for regression problems, is used. $comp$ quantifies the complexity of trees and penalizes for more complex models in the tree selection process. In CART, the complexity is a non-negative function of the terminal-nodes. Note that finding $\hat{\theta}$ requires a search over all $\Theta_m$.

@@

\section{Implementation and application in practice}
\label{impl}

\code{evtree} provides an efficient implementation of an evolutionary algorithm that builds classification trees in \proglang{R}. CPU- and memory- intensive tasks are fully computed in \proglang{C++}, while the user interfaces and plot functions are written in \proglang{R}. The \code{.C()} interface \citep{chambers2008software} was used to pass arguments between the two languages. \code{evtree} depends on the \pkg{partykit} package  \citep{hothorn2009partykit}, which provides an infrastructure for representing, summarizing, and visualizing tree-structured models. The supplied functionality includes standard \code{print()}, \code{summary()}, and \code{plot()} functions to display trees and a \code{predict()} function to predict the class labels of new data. The arguments of  \code{evtree()} are
%
\begin{Code}
  evtree(formula, data = list(), weights = NULL, subset = NULL,
    control = evtree.control(...), ...)
\end{Code}
%
where \code{formula} and \code{data} are a standard \proglang{R} formula of type \code{y ~ x1 + x2} with the respective data. \code{weights} is an optional vector specifying weights to be used in the fitting process. The option vector subset specifies a subset of observations to be used in the fitting process. \code{control} comprises a list of control parameters
%
\begin{Code}
  evtree.control(minbucket = 7L, minsplit = 20L, maxdepth = 9L, 
    niterations = 10000L, ntrees = 100L, alpha = 1,
    operatorprob = list(pmutatemajor = 0.2, pmutateminor = 0.2,
      pcrossover = 0.2, psplit = 0.2, pprune = 0.2),
    seed = NULL, ...)
\end{Code}
%
where the parameters \code{minbucket}, \code{minsplit}, and \code{maxdepth} constrain the solution to a minimum number of observations in each terminal node, a minimum number of observation in each internal node and a maximum tree depth. Note that the memory requirements increase by the square of the maximum tree depth. Parameter \code{alpha} regulates the complexity part of the evaluation function and refers to $\alpha$ in Equation~\ref{eq:classification} for classification trees and Equation~\ref{eq:regression} for regression trees. \code{niterations} and \code{ntrees} specify the maximum number of iterations and the number of trees in the population. With the argument \code{operatorprob}, user-specified probabilities for the variation operators can be defined. Argument \code{seed} is an optional seed for the random number generator, which allows models to be reproduced by re-using the same seed. If not specified, the random number generator is initialized by the system time. In this case, iterative calls of \code{evtree} will possibly result in different models.

An interesting application for classification tree analysis is target marketing, where limited resources are aimed at a distinct group of potential customers. An example of a marketing case study is provided by the \textit{Bookbinder's Book Club} \citep{lilien2004} dataset. The data is a marketing case study about a (fictitious) American book club. In this case study, a brochure of the book \textit{The Art History of Florence} was sent to 20,000 customers and 1,806 of which bought the book. The dataset contains a subsample of 1,300 customers for building a predictive model for the customers choice of buying the advertised book. Besides predictive accuracy, model complexity is a crucial issue in this application. More simple trees are easier to interpret and communicable to marketing experts and management professionals.

In the following analysis of the \textit{Bookbinder's Book Club} data \code{evtree} is used in conjunction with \code{rpart} and \code{ctree}. All trees are constrained to have a minimum number of 26 observations per terminal node ($2\%$ of the total number of observations). To avoid excessively large trees \code{ctree} is called with $\alpha = 1\%$. To provide uniform visualizations and predictions of the fitted models the \code{as.party()} function is used to convert \pkg{rpart} objects into objects of class \pkg{party}. The sample session starts with the construction and visualization of trees using the \code{rpart} and \code{ctree} methods. 
%
<<recursive-trees>>=
data("BBBClub", package = "evtree")
library("rpart")
set.seed(1090)
rp <- as.party(rpart(choice ~ ., data = BBBClub, minbucket = 26))
ct <- ctree(choice ~ ., data = BBBClub, mincrit = 0.99, minbucket = 26)
plot(rp)
plot(ct)
@
%
The resulting trees are displayed in Figure~\ref{fig:example_rpart_ctree}.
%
\setkeys{Gin}{width=1\textwidth}
\begin{figure}[t!]
\centering
\begin{minipage}{1\textwidth}
\begin{flushleft} \code{rpart} \end{flushleft}
\centering
<<rp-plot, fig=TRUE, echo=FALSE, width=12, height=6.8>>=
plot(rp)
@
\end{minipage}
\begin{minipage}{1\textwidth}
\begin{flushleft} \code{ctree} \end{flushleft}
\centering
<<ct-plot, fig=TRUE, echo=FALSE,  width=12, height=6.8>>=
plot(ct)
@
\end{minipage}
\caption{\label{fig:example_rpart_ctree} Trees constructed by \code{rpart} (upper panel) and \code{ctree} (lower panel). Target variable is the customers choice of buying the book. The variables used for splitting are: \code{art} number of art books purchased; \code{first} number of months since the first purchase; \code{freq} number of books purchased at the BBB Club; \code{gender} the costumers gender.
}
\end{figure}
%
With the objective of building a smaller, but at least equally accurate tree, \code{evtree} is constrained to a maximum tree depth of $2$. The resulting tree is displayed in Figure~\ref{fig:example_evtree}. Alternatively the tree can be examined by the textual output provided by the \code{print()} function.
%
<<evtree-plot>>=
ev <- evtree(choice ~ ., data = BBBClub, minbucket = 26, maxdepth = 2)
plot(ev)
print(ev)
@
%
\setkeys{Gin}{width=1\textwidth}
\begin{figure}[t!]
\begin{flushleft} \code{evtree} \end{flushleft}
\centering
<<iris-plot, fig=TRUE, echo=FALSE, width=12, height=7>>=
plot(ev)
@
\caption{\label{fig:example_evtree} Tree constructed by \code{evtree}. Target variable is the customers choice of buying the book. The variables used for splitting are: \code{art} number of art books purchased; \code{first} number of months since the first purchase}
\end{figure}
%
While the \code{rpart} model has $5$ terminal nodes and a tree depth of $4$, the tree constructed by \code{ctree} has $7$ terminal nodes and a tree depth of $3$. With a tree depth of $2$ and $4$ terminal nodes the \code{evtree} model is much simpler. 
%
<<evtree-performance>>=
## accuracy: misclassification rate
mc <- function(obj) 1 - mean(predict(obj) == BBBClub$choice)
c("evtree" = mc(ev), "rpart" = mc(rp), "ctree" = mc(ct))

## complexity: number of terminal nodes
c("evtree" = width(ev), "rpart" = width(rp), "ctree" = width(ct))
@
The accuracy of the three models are similar. With a misclassification rate of $24.3\%$, the \code{evtree} model is slightly worse than the \code{rpart} model ($23.8\%$) and better than the \code{ctree} model ($24.9\%$). 
%
<<evtree-structure>>=
## compare structure of predictions
ftable(tab <- table(evtree = predict(ev), rpart  = predict(rp),
  ctree  = predict(ct), observed = BBBClub$choice))

## compare customer predictions only (absolute, proportion correct)
sapply(c("evtree", "rpart", "ctree"), function(nam) {
  mt <- margin.table(tab, c(match(nam, names(dimnames(tab))), 4))
  c(abs = as.vector(rowSums(mt))[2],
    rel = round(100 * prop.table(mt, 1)[2, 2], digits = 3))
})
@
%
Although the constructed trees are considerably different, the structure of the predictions are relatively similar. In this case, \code{evtree} classifies less customers ($186$) as buyers as \code{rpart} ($216$) and \code{ctree} ($278$). Of the customers classified as buyers, $72.6\%$ (\code{evtree}), $70.8\%$ (\code{rpart}) and $63.7\%$ (\code{ctree}) actually bought the book. 
%
<<evtree-performance>>=
rp <- as.party(rpart(choice ~ ., data = BBBClub, minbucket = 26, 
maxdepth = 2))
ct <- ctree(choice ~ ., data = BBBClub, mincrit = 0.99, minbucket = 26, 
maxdepth = 2)
## accuracy: misclassification rate
mc <- function(obj) 1 - mean(predict(obj) == BBBClub$choice)
c("evtree" = mc(ev), "rpart" = mc(rp), "ctree" = mc(ct))
## complexity: number of terminal nodes
c("evtree" = width(ev), "rpart" = width(rp), "ctree" = width(ct))
@
%
\code{rpart} and \code{ctree} models are fitted with the same constrained on the tree's depth (\code{maxdepth = 2}) as the previous \code{evtree} model. With a misclassification rate of $24.3\%$, the \code{evtree} model is more accurate than the \code{rpart} model ($26.2\%$) and the \code{ctree} model ($25.5\%$). This shows an particular advantage of \code{evtree}, which is its ability to built the best tree for a given tree size.
